---
title: "Gene clustering across multiple sample groups"
author: "Jim Zhang"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_caption: yes
    toc: yes
---

```{r global_setup, include=FALSE}
knitr::opts_chunk$set(dpi=300, fig.pos="H", fig.width=8, fig.height=6, dev=c('png', 'pdf'), echo=FALSE, warning=FALSE, message=FALSE);

library(awsomics);
library(gplots);
library(knitr);
library(rmarkdown); 

# fn.yaml<-"~/R/source/DEGandMore/examples/MultiGroupCluster/ClReport.yml"; # comment it out if running through knitr::knit function
# fn.yaml<-"/nas/is1/zhangz/projects/barr/2015-11_Rat_Brain_Development/result/gene_clustering/Juvenile_SC_injury/ClReport.yml"; 

# Loading inputs from yaml file
# if (!exists('fn.yaml')) stop('Input file not found\n'); 
# writeLines(yaml::as.yaml(yml), fn.yaml);
# yml <- yaml::yaml.load_file(fn.yaml);  

expr<-readRDS(yml$input$data); 
anno<-readRDS(yml$input$annotation);
smpl<-readRDS(yml$input$sample);
gset<-readRDS(yml$input$geneset); 

if (!setequal(rownames(expr), rownames(anno))) stop("Data matrix and gene annotation have different row names.\n");
if (!setequal(colnames(expr), rownames(smpl))) stop("Data matrix and sample metadata have different names.\n");
expr<-expr[rownames(anno), rownames(smpl)]; 

prms<-yml$parameter;
```

**_[Go back to project home](`r yml$project$Home`)_**

# Project background

```{r project, include=FALSE}
lns<-lapply(names(yml$project), function(nm) {
  c(paste('##', nm), '\n', yml$project[[nm]], '\n'); 
});
lns<-paste(do.call('c', lns), collapse='\n'); 
```

`r lns`

**_[Go back to project home](`r yml$project$Home`)_**

# Samples and methods

```{r methods, include=FALSE}
CreateDatatable(smpl, paste(yml$output, 'samples.html', sep='/'));
lns<-lapply(names(yml$methods), function(nm) {
  c(paste('##', nm), '\n', yml$methods[[nm]], '\n'); 
});
lns<-paste(do.call('c', lns), collapse='\n'); 
```

## Samples

There are a total of `r ncol(expr)` samples. Click [here](samples.html) to view full table of samples

`r lns`


# Analysis and results

**_[Go back to project home](`r yml$project$Home`)_**

## ANOVA analysis of each gene across sample

```{r anova, include=FALSE}
f<-as.factor(smpl[, prms$term[1]]); 
l<-unique(smpl[, prms$term[1]]);
grp<-split(rownames(smpl), as.vector(f))[l];
m<-sapply(grp, function(s) rowMeans(expr[, s, drop=FALSE])); 
mn<-apply(expr, 1, min);
mx<-apply(expr, 1, max); 

p<-apply(expr, 1, function(d) summary(aov(d ~ f))[[1]][1, 5]); 
q<-p.adjust(p, method='BH');
stat<-cbind(m, Min=mn, Max=mx, Range=mx-mn, pANOVA=p, FDR=q); 

CreateDatatable(cbind(anno, FormatNumeric(stat)), paste(yml$output, 'anova.html', sep='/'), caption="ANOVA results")
```

We applied ANOVA analysis on each gene to test its differential expression across samples of different groups.

  - There are `r nrow(expr)` total genes
  - There are `r ncol(expr)` total samples
  - There are `r length(l)` sample groups: `r paste(l, collapse=', ')`

Click [here](anova.html) to view full ANOVA result table. 

**_[Go back to project home](`r yml$project$Home`)_**

## Gene clustering

### Select genes differentially expressed across groups

Select genes using the following steps:

  1. Select genes with FDR less than `r prms$selection$fdr`
  2. Stop if less than `r prms$selection$min` genes were left; otherwise, select those with ANOVA p values less than `r prms$selection$p`
  3. Stop if less than `r prms$selection$min` genes were left; otherwise, select those with range (_max-min_) greater than `r prms$selection$range`
  4. If there are still more than `r prms$selection$max` genes left, select the top `r prms$selection$max` with the biggest ranges
  
```{r selection, include=FALSE}
e<-expr[q<=prms$selection$fdr, , drop=FALSE]; 
if (nrow(e) > prms$selection$min) { # continue if there are more selected genes than minimum
  s<-stat[rownames(e), , drop=FALSE];
  s<-s[order(s[, 'pANOVA']), , drop=FALSE];
  e<-e[rownames(s), , drop=FALSE]; 
  e<-e[1:max(prms$selection$min, max(which(s[, 'pANOVA']<=prms$selection$p))), , drop=FALSE]; 
  
  if (nrow(e) > prms$selection$min) { # continue if there are more selected genes than minimum 
    s<-stat[rownames(e), , drop=FALSE];
    s<-s[rev(order(s[, 'Range'])), , drop=FALSE];
    e<-e[rownames(s), , drop=FALSE]; 
    e<-e[1:max(prms$selection$min, max(which(s[, 'Range']>=prms$selection$range))), , drop=FALSE]; 
  }
  
  if (nrow(e) > prms$selection$max) e<-e[1:prms$selection$max, , drop=FALSE]; # trim if more selected genes than maximum
}
```

A total of **`r nrow(e)`** genes were selected. The genes were used for the next step to identify gene clusters

**Figure** Hierarchical clustering of samples using all genes.

```{r hc_all, include=TRUE, out.width='800px'}
plot(hclust(as.dist(1-cor(expr))), main='Sample clustering using all genes', xlab='Sample', sub='');
```

**Figure** Hierarchical clustering of samples using just selected genes.

```{r hc_clustered, include=TRUE, out.width='800px'}
plot(hclust(as.dist(1-cor(e))), main='Sample clustering using selected significant genes', xlab='Sample', sub='');
```

### Cluster selected genes

Genes selected by the last step were clustering using the following steps:

  - Create a hierchical tree based on gene-gene correlation
  - Cut the tree at height `r prms$cluster$height`, which will classify genes into clusters. Then apply the following steps to refine the clusters
    - Calculate correlation of each gene to the centroid (median) of its cluster. Remove the genes if the correlation is less than `r prms$cluster$corr`
    - Remove clusters with size less than `r 100*prms$cluster$size`% of the expected size (the expected size is 50 if there are 500 genes and 10 clusters)
    - If there are less than `r length(grp)` (the number of sample groups) clusters left, reduce the height cutoff by 0.05 and repeat this step until there are enough clusters
  - Merge the 2 most similar clusters if the correlation of their centroids is greater than or equal to `r prms$cluster$corr`. Repeat this step until no 2 clusters are that similar

```{r cluster_seeding, include=FALSE}
h<-prms$cluster$height;
sz<-prms$cluster$size; # minimal size of a cluster, ratio against expected size (ex. if size=0.2, minimal cluster size is 0.2*500/10 when there are 500 genes and 10 clusters in total)

# Hierarchical clustering
hc<-hclust(as.dist(1-cor(t(e)))); 
ct<-cutree(hc, h=h);
cl<-split(names(ct), ct); 
cl<-sapply(cl, function(c) {
  x<-e[c, ]; 
  md<-apply(x, 2, median); 
  r<-as.vector(cor(md, t(x))[1, ]); # remove outliers 
  c[r>=prms$cluster$corr]
}); 
cl<-cl[sapply(cl, length)>=sz*nrow(e)/length(cl)];

# Reduce cutoff if number of clusters is less than number of groups
while(length(cl) < length(grp)) {
    h<-h-0.05;
    ct<-cutree(hc, h=h);
    cl<-split(names(ct), ct); 
    cl<-sapply(cl, function(c) {
    x<-e[c, ]; 
    md<-apply(x, 2, median); 
    r<-as.vector(cor(md, t(x))[1, ]); # remove outliers 
    c[r>=prms$cluster$corr]
  }); 
  cl<-cl[sapply(cl, length)>=sz*nrow(e)/length(cl)];
}

# merge similar clusters
flag<-TRUE;
while(flag) {
  cat('Number of clusters ', length(cl), '\n'); 
  ms<-sapply(cl, function(cl) colMeans(expr[cl, ])); 
  tr<-cutree(hclust(as.dist(1-cor(ms))), k=length(cl)-1); # find the 2 most similar clusters
  i<-tr[duplicated(tr)];
  c<-ms[, tr==i]; 
  r<-cor(c[, 1], c[, 2]); 
  p<-cor.test(c[, 1], c[, 2])$p.value[[1]]; 
  if (r>prms$cluster$merge$corr & p<prms$cluster$merge$p) {
    cl[tr==i][[1]]<-as.vector(unlist(cl[tr==i]));
    cl<-cl[names(cl)!=names(i)]; 
  } else flag<-FALSE;
}

# Sort clusters
m<-sapply(cl, function(cl) colMeans(expr[cl, ])); 
#m<-sapply(grp, function(g) colMeans(m[g, ])); 
ind<-apply(m, 2, function(x) which(x==max(x)));  
cl<-cl[order(ind)]; 
names(cl)<-paste('Cluster', 1:length(cl), sep='_'); 

# preparing for the next blocks
ms<-sapply(cl, function(cl) colMeans(expr[cl, ])); 
ms<-data.frame(t(ms));
colnames(ms)<-colnames(e); 
rownames(ms)<-paste(rownames(ms), ' (n=', sapply(cl, length),  ')', sep=''); 

sortColumn<-function(ms, g) {
  corr<-as.vector(cor(rowMeans(ms[, g[[2]], drop=FALSE]), ms[, g[[1]], drop=FALSE]));
  g[[1]]<-g[[1]][order(corr)];
  for (i in 2:length(g)) {
    corr<-as.vector(cor(rowMeans(ms[, g[[i-1]], drop=FALSE]), ms[, grp[[i]], drop=FALSE]));
    g[[i]]<-g[[i]][rev(order(corr))]; 
  }
  ms[, as.vector(unlist(g))]; 
}

n<-sapply(cl, length); 
d<-expr[stat[, 'pANOVA']<=prms$recluster$p, , drop=FALSE]; 
```

The initial analysis identified `r length(cl)` gene clusters using just selected significant genes. A total of **`r sum(n)`** genes were clustered. 

**Figure** This plot shows below the average expression levels of each cluster. Values indicate number of standard deviation from mean of all samples. 

```{r plot_cluster_mean, include=TRUE, fig.width=CalculateColoredBlockSize(ms)[1], fig.height=CalculateColoredBlockSize(ms)[2], out.width='600px'}
if (prms$plot$rescale) x<-t(scale(t(ms))) else x<-ms;
PlotColoredBlock(sortColumn(x, grp), key='Average expression', groups=grp); 
```

### Refine clusters

We further refine the gene clusters with the following steps: 

  - Select all `r nrow(d)` genes with p values less than `r prms$recluster$p` to continue
  - Assign selected genes to clusters:
    - Calculate centroid (median expression level of all genes in the cluster) of each cluster
    - Calculate correlation coefficient of each gene to centroid of each cluster to get a `r nrow(d)` X `r length(cl)` matrix
    - Assign a gene to a cluster if its correlation coefficient to the cluster is greater than `r prms$recluster$r`, and the correlation coefficient to any other cluster is at least `r prms$recluster$diff` less
    - Repeat this reclustering steps for `r prms$recluster$times` times unless the reclustering converged

```{r cluster_refine, include=FALSE}
reCl<-function(d, cl, r, dif) {
  md<-sapply(cl, function(cl) apply(d[cl[cl %in% rownames(d)], , drop=FALSE], 2, median));
  corr<-cor(t(d), md); 
  c<-lapply(1:ncol(corr), function(i) {
    mx<-apply(corr[, -i, drop=FALSE], 1, max);
    rownames(corr)[corr[, i]>=r & (corr[, i]-mx)>dif]; 
  });
  c;
}

cls<-list(cl, reCl(d, cl, prms$recluster$r, prms$recluster$diff)); 

while (!identical(cls[[length(cls)]], cls[[length(cls)-1]]) & length(cls)<=prms$recluster$times) {
  cls[[length(cls)+1]]<-reCl(d, cls[[length(cls)]], prms$recluster$r, prms$recluster$diff); 
  cat("Reclustering #", length(cls)-1, '\n'); 
}

cl<-cls[[length(cls)]]; 
names(cl)<-paste('Cluster', 1:length(cl), sep='_'); 

# summary cluster sizes
n<-sapply(cls, function(c) sapply(c, length)); 
n<-data.frame(n); 
colnames(n)<-paste('Cycle #', 0:(ncol(n)-1), sep=''); 

ms<-sapply(cl, function(cl) colMeans(expr[cl, ])); 
ms<-data.frame(t(ms));
colnames(ms)<-colnames(expr); 
rownames(ms)<-paste(rownames(ms), ' (n=', sapply(cl, length),  ')', sep=''); 

if(identical(cls[[length(cls)]], cls[[length(cls)-1]])) msg<-paste("The reclustering converged after", length(cls)-1, 'cycles.') else msg<-paste("The reclustering didn't converge after", length(cls)-1, 'cycles.')
```

`r msg`. A total of **`r sum(sapply(cls[[length(cls)]], length))`** genes were clustered. 

**Figure** This plot shows below the average expression levels of each cluster after `r ncol(n)-1` cycles of reclustering. Values indicate number of standard deviation from mean of all samples. 

```{r plot_recluster_mean, include=TRUE, fig.width=CalculateColoredBlockSize(ms, 1, 6)[1], fig.height=CalculateColoredBlockSize(ms, 1, 6)[2], out.width='600px'}
if (prms$plot$rescale) x<-t(scale(t(ms))) else x<-ms;
PlotColoredBlock(sortColumn(x, grp), key='Average expression', groups=grp); 
```

```{r summary_clusters, include=FALSE}
ids<-as.vector(unlist(cls[[length(cls)]])); 
c<-rep(names(cl), sapply(cl, length)); 
tbl1<-cbind(Cluster=c, FormatNumeric(expr[ids, ]), anno[ids, ]); 
tbl2<-cbind(Cluster=c, FormatNumeric(stat[ids, ]), anno[ids, ]); 

fn1<-CreateDatatable(tbl1, paste(yml$output, 'clustered_data.html', sep='/'), caption="Expression data of clustered genes");
fn2<-CreateDatatable(tbl2, paste(yml$output, 'clustered_stat.html', sep='/'), caption="ANOVA statistical results of clustered genes"); 

write.csv2(tbl1, paste(yml$output, 'clustered_data.csv', sep='/')); 
write.csv2(tbl2, paste(yml$output, 'clustered_stat.csv', sep='/')); 
write.csv2(cbind(expr, anno), paste(yml$output, 'all_genes_data.csv', sep='/')); 
write.csv2(cbind(stat, anno), paste(yml$output, 'all_genes_stat.csv', sep='/')); 

saveRDS(tbl1, paste(yml$output, 'clustered_data.rds', sep='/')); 
saveRDS(tbl2, paste(yml$output, 'clustered_stat.rds', sep='/')); 
saveRDS(cbind(expr, anno), paste(yml$output, 'all_genes_data.rds', sep='/')); 
saveRDS(cbind(stat, anno), paste(yml$output, 'all_genes_stat.rds', sep='/')); 

saveRDS(gset, paste(yml$output, 'all_genesets.rds', sep='/')); 
```

Result tables:

  - Normalized expressiond data of clustered genes
    - Click [here](clustered_data.html) to view online
    - Click [here](clustered_data.csv) to download table
  - Statistical results of clustered genes
    - Click [here](clustered_stat.html) to view online
    - Click [here](clustered_stat.csv) to download table

**_[Go back to project home](`r yml$project$Home`)_**

## Analysis of individual clusters

```{r cluster_detail, include=FALSE}
fn.temp<-paste(yml$output, 'ClDetail.Rmd', sep='/'); 
if (yml$input$remote) {
  if (!RCurl::url.exists(yml$input$subtemplate)) stop("Template Rmd file ', yml$input$subtemplate, ' not exists\n");
  writeLines(RCurl::getURL(yml$input$subtemplate)[[1]], fn.temp);
} else {
  file.copy(yml$input$subtemplate, fn.temp); 
}

fns<-sapply(1:length(cl), function(i) {
  mtrx<-expr[cl[[i]], , drop=FALSE]; 
  name<-names(cl)[i]; 
  grps<-grp;
  univ<-rownames(anno); 
  path<-paste(yml$output, name, sep='/'); 
  home<-"../index.html";
  
  if (prms$plot$zero) mtrx<-mtrx-rowMeans(mtrx[, grp[[1]], drop=FALSE]); 
  
  if (!file.exists(path)) dir.create(path, recursive = TRUE); 

  fn.html<-paste(name, '.html', sep=''); 
  
  rmarkdown::render(fn.temp, output_file=fn.html, output_dir=paste(yml$output, name, sep='/'), quiet=TRUE); 
}); 

```

```{r cluster_link, include=FALSE}
lnk<-paste(names(cl), paste(names(cl), '.html', sep=''), sep='/'); 
lns<-paste('  - [', names(cl), '](', lnk, ')', sep='');
lns<-paste(lns, collapse='\n'); 
```

Click cluster names to see more detaied information about individual clusters

`r lns`

**_[Go back to project home](yml$project$Home)_**

## Extra plots

**Figure** Plot the group means and standard errors of all cluster as data series 

```{r plot_series, include=TRUE, out.width='750px'}
ms<-sapply(cl, function(cl) colMeans(expr[cl, ]));
if (prms$plot$rescale) ms<-t(scale(t(ms))); 
m<-sapply(grp, function(g) colMeans(ms[g, , drop=FALSE]));
se<-sapply(grp, function(g) apply(ms[g, , drop=FALSE],  2, function(x) sd(x)/sqrt(length(x)))); 
if (prms$plot$zero) m<-m-m[, 1];
PlotSeries(m, se, c('', prms$plot$ylab)); 
```

**Figure** Heatmap of group and cluster means

```{r plot_means, include=TRUE, out.width='600px'}
ms<-sapply(cl, function(cl) colMeans(expr[cl, ]));
if (prms$plot$rescale) ms<-t(scale(t(ms))); 
m<-sapply(grp, function(g) colMeans(ms[g, , drop=FALSE]));
colnames(m)<-names(grp);
rownames(m)<-names(cl); 
PlotColoredBlock(m, key='Cluster average'); 
```

**Figure** Group means of all clustered genes

```{r plot_group_means, include=TRUE, out.width='600px'}
d<-expr[unlist(cl, use.names = FALSE), , drop=FALSE];
d<-sapply(grp, function(g) rowMeans(d[, g, drop=FALSE])); 
colnames(d)<-names(grp); 
PlotColoredBlock(d, num.breaks = 127, groups=grp, key='Group average'); 
abline(h=c(nrow(d), nrow(d)-cumsum(sapply(cl, length)), lwd=1)); 
```

**Figure** All samples and all clustered genes

```{r plot_heatmap, include=TRUE, out.width='900px', out.height='1200px'}
d<-expr[unlist(cl, use.names = FALSE), , drop=FALSE];
d<-sortColumn(d, grp); 
rownames(d)<-paste(rownames(d), anno[rownames(d), 1], sep=':'); 
if (prms$plot$rescale) d<-t(scale(t(d))); 
PlotColoredBlock(d, num.breaks = 127, groups=grp, key='Expression level'); 
abline(h=c(nrow(d), nrow(d)-cumsum(sapply(cl, length)), lwd=1)); 
abline(v=c(0, cumsum(sapply(grp, length))), lwd=1); 
```

**Figure** This plot traces the change of cluster sizes after each reclustering cycle. 

```{r plot_recluster_size, include=TRUE, fig.width=CalculateColoredBlockSize(n, 1, 6)[1], fig.height=CalculateColoredBlockSize(n, 1, 6)[2], out.width='750px'}
rownames(n)<-paste(rownames(n), ' (n=', n[, ncol(n)], ')', sep='')
PlotColoredBlock(n, key='Cluster size'); 
```

```{r wrap_up, eval=FALSE, include=FALSE}
x<-strsplit(yml$output, '/')[[1]]; 
fn.zip<-paste(x[length(x)], '.zip', sep='');
fn.zip.fn<-paste(yml$output, fn.zip, sep='/');
if (file.exists(fn.zip.fn)) file.remove(fn.zip.fn); 
zip(fn.zip, yml$output, "-rj9X", zip='zip'); 
file.rename(fn.zip, fn.zip.fn); 
```

**_[Go back to project home](`r yml$project$Home`)_**

***
_END OF DOCUMENT_